---
title: "ggplot2 and palmsplusr"
author:
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4-ggplot2-palmsplusr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Install ggplot2

The development version of ggplot2 contains spatial plotting functions so it needs to be installed from Github. You will need to install devtools if you haven't already.

```{r, eval=FALSE}
devtools::install_github("tidyverse/ggplot2")
```


## Plotting 

The `geom_sf` is a new ggplot2 geom specifically for plotting simple features. You can plot points, lines, and polygons. Here is an example plotting one participant's trajcetories, school and home on a map.

```{r message=FALSE, warning=FALSE}
library(palmsplusr)
library(ggplot2)

data(list = c("palms", "home", "school"))
home.buffer <- palms_buffer(home, 100)

# Load default fields
palms_remove_tables()
palms_load_defaults(palms_epoch(palms))

# Build palmsplus and trajectories
palmsplus <- palms_build_palmsplus(palms)
trajectories <- palms_build_trajectories(palmsplus)

# Plot
ggplot() +
  geom_sf(data=trajectories, aes(colour = factor(mot)), lwd = 0.9) +
  geom_sf(data=home.buffer, aes(fill="h"), alpha = 0.8) +
  geom_sf(data=school, aes(fill="s"), alpha = 0.8) +
  scale_fill_manual(labels = c("Home (100m)", "Schoolyard"),
                    values = c("red", "blue")) +
  scale_color_manual(breaks = c("1", "2", "3"),
                     labels = c("Walk", "Bicycle", "Vehicle"),
                     values = c("1"="forestgreen","2"="lightblue","3"="orange")) +
  theme_minimal() +
  labs(colour = "Trajectories", fill = "Polygons")
```

This is an excellent way to visualise and debug your datasets after they have been built. It is easy it loop through all participants and save their plots to file.

```{r, eval=FALSE}
# Loop through all unique identifiers
for (i in unique(trajectories$identifier)) {
  
  # Get geometry for this person only
  t <- trajectories %>% filter(identifier == i)
  h <- home.buffer %>% filter(identifier == i)
  s <- school %>% filter(school_id == as.numeric(participant_basis[participant_basis$identifier == i, "school_id"]))

  # Plot the 3 geometry layers + add colours and title
  p <- ggplot() +
    geom_sf(data = t, aes(colour = factor(mot))) +
    geom_sf(data = h, aes(fill = "h"), alpha = 0.8) +
    geom_sf(data = s, aes(fill = "s"), alpha = 0.8) +
    scale_fill_manual(labels = c("Home (100m)", "Schoolyard"),
                      values = c("red", "blue")) +
    scale_color_manual(breaks = c("1", "2", "3"),
                       labels = c("Walk", "Bicycle", "Vehicle"),
                       values = c("1"="forestgreen","2"="lightblue","3"="orange")) +
    theme_minimal() +
    labs(title = i, colour = "Trajectories", fill = "Polygons")

  # Save the plot to file
  ggsave(file = file.path("plots", paste0(i, ".png")), plot = p, width = 18, 
         height = 18, units = "cm")
}
```
