---
title: "A complete example"
author:
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2-complete-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article will demonstrate a real world example with a 197-participant dataset.

## Reading data

It is wise to read in all data that you will need at the beginning. This includes the PALMS dataset, and any other shapefiles and csv files. Any geoprocessing, such as buffering can also be done here.

```{r message=FALSE, warning=FALSE}
library(palmsplusr)
library(readr)

# Load PALMS dataset
palms <- read_palms("D:/data/csv/palms_output.csv")

# Load other csvs
participant_basis <- read_csv("D:/data/csv/participant_basis.csv")
class_timetable <- read_csv("D:/data/csv/class_timetable.csv")

# Load shapefiles
home <- read_sf("D:/data/shapefiles/home.shp")
home <- home %>% rename(identifier = Identifier)
school <- read_sf("D:/data/shapefiles/school.shp")
school <- school %>% rename(school_id = schoolID)

# Buffer home points
home.buffer <- palms_buffer(point = home, distance = 100, crs = 2193)
```

The participant_basis file is used to link participant identifier number with the correct school and classroom: 
```{r}
head(participant_basis, n = 5)
```

The class_timetable file details start and end times for school and recess preiods, for each school/classroom:
```{r}
head(class_timetable, n = 5)
```

## Creating field, domain, and trajectory location tables

This section is the bulk of the setup. All formulas for building `palmsplus`, `days`, `trajectories`, and `multimodal` are created. 

Notice that `palms_load_defaults()` is used as the base, and then extra fields are then appended:

```{r}
# Remove all field tables that exist
palms_remove_tables()

# load defaults for:
# * palmsplus_fields
# * trajectory_fields
# * multimodal_fields
epoch <- palms_epoch(palms)
palms_load_defaults(epoch)

# palmsplus fields (from other files)
palms_add_field("in_school_time", "palms_in_time(., i, class_timetable, participant_basis, school_start, school_end)")
palms_add_field("in_recess1", "palms_in_time(., i, class_timetable, participant_basis, recess1_start, recess1_end)")
palms_add_field("in_recess2", "palms_in_time(., i, class_timetable, participant_basis, recess2_start, recess2_end)")
palms_add_field("at_home", "palms_in_polygon(., filter(home.buffer, identifier == i), identifier)")
palms_add_field("at_school", "palms_in_polygon(., filter(school, school_id == participant_basis %>% filter(identifier == i) %>% pull(school_id)))")

# days domains
palms_add_domain("home", "at_home")
palms_add_domain("school", "(!at_home & at_school & in_school_time)")
palms_add_domain("transport", "!at_home & !(at_school & in_school_time) & (pedestrian | bicycle | vehicle)")
palms_add_domain("leisure", "!at_home & !(at_school & in_school_time) & !pedestrian & !bicycle & !vehicle")

# trajectory locations
palms_add_trajectory_location("home_school", "at_home", "at_school")
palms_add_trajectory_location("school_home", "at_school", "at_home")
palms_add_trajectory_location("home_home", "at_home", "at_home")
palms_add_trajectory_location("school_school", "at_school", "at_school")
```

Note that in **palmsplusr** there are five field tables:

* `palmsplus_fields`
* `palmsplus_domains`
* `trajectory_fields`
* `trajectory_locations`
* `multimodal_fields`

We didn't specify any additional `trajectory_fields` or `multimodal_fields` as the default fields suffice.

## Build palmsplus, days, and trajectories

Once all of the formula tables are created, it is very simple to build palmsplus, and then create days and trajectories (and multimodal trips).

```{r}
# Start the timer
t <- proc.time()

# Build palmsplus
palmsplus <- palms_build_palmsplus(palms)

# Build days
days <- palms_build_days(palmsplus)

# Build trajectories
trajectories <- palms_build_trajectories(palmsplus)

# Build multimodal trips
multimodal <- palms_build_multimodal(trajectories, 200, 10)

cat("Run time:", (proc.time() - t)[[1]]/60, "minutes")
```

## Saving results

```{r, eval=FALSE}
write_csv(palmsplus, "palmsplus.csv")
write_csv(days, "days.csv")
write_csv(trajectories, "trajectories.csv")
write_csv(trajectories, "multimodal.csv")

st_write(palmsplus, "palmsplus.shp")
st_write(trajectories, "trajecories.shp")
st_write(trajectories, "multimodal.shp")
```


