---
title: "A complete example"
author:
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2-complete-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article will demonstrate a real world example with a 191-participant dataset.

Note: this uses real local data, so the vignette is not actually built (only used for online documentation).

## Reading data

All data should be loaded at the beginning of the script. This includes the PALMS dataset, and any other shapefiles and csv files. Any preprocessing, such as buffering can also be done here.

```{r message=FALSE, warning=FALSE}
library(palmsplusr)
library(readr)

# Load PALMS dataset
palms <- read_palms("D:/data/csv/palms_output.csv")

# Load other csvs
participant_basis <- read_csv("D:/data/csv/participant_basis.csv")
class_timetable <- read_csv("D:/data/csv/class_timetable.csv")

# Load shapefiles (home points & schoolyard polygons)
home <- read_sf("D:/data/shapefiles/home.shp")
school <- read_sf("D:/data/shapefiles/school.shp")

# Buffer home points
home.buffer <- palms_buffer(point = home, distance = 100, crs = 2193)
```

The participant_basis file is used to link participant identifier with the correct school and classroom: 
```{r}
head(participant_basis, n = 5)
```

The class_timetable file details start and end times for school and recess preiods, for each school/classroom combination:
```{r}
head(class_timetable, n = 5)
```

## Creating the five field tables

All formulas for building the `palmsplus`, `days`, `trajectories`, and `multimodal` datasets are created. This is done by creating the five field tables:

* `palmsplus_fields`
* `trajectory_fields`
* `multimodal_fields`
* `palmsplus_domains`
* `trajectory_locations`


The first three of these tables can be populated using the `palms_load_defaults()`. I use this function as a base, but then add additional fields that are specific to my dataset and research questions. To see a list of fields that are included by default, please see [the building formulas article](https://thets.github.io/palmsplusr/articles/article-3-building-formulas.html).

```{r}
# Remove all field tables that exist
palms_remove_tables()

# Load defaults
palms_load_defaults(palms_epoch(palms))

# palmsplus_fields (from other files)
palms_add_field("in_school_time", "palms_in_time(., i, class_timetable, participant_basis, school_start, school_end)")
palms_add_field("in_recess1", "palms_in_time(., i, class_timetable, participant_basis, recess1_start, recess1_end)")
palms_add_field("in_recess2", "palms_in_time(., i, class_timetable, participant_basis, recess2_start, recess2_end)")
palms_add_field("at_home", "palms_in_polygon(., filter(home.buffer, identifier == i), identifier)")
palms_add_field("at_school", "palms_in_polygon(., filter(school, school_id == participant_basis %>% filter(identifier == i) %>% pull(school_id)))")

# palmsplus_domains
palms_add_domain("home", "at_home")
palms_add_domain("school", "(!at_home & at_school & in_school_time)")
palms_add_domain("transport", "!at_home & !(at_school & in_school_time) & (pedestrian | bicycle | vehicle)")
palms_add_domain("leisure", "!at_home & !(at_school & in_school_time) & !pedestrian & !bicycle & !vehicle")

# trajectory_locations
palms_add_trajectory_location("home_school", "at_home", "at_school")
palms_add_trajectory_location("school_home", "at_school", "at_home")
palms_add_trajectory_location("home_home", "at_home", "at_home")
palms_add_trajectory_location("school_school", "at_school", "at_school")
```

I didn't specify any additional `trajectory_fields` or `multimodal_fields` as the default fields will suffice. 

## Build palmsplus, days, trajectories, and multimodal

Once all of the formula tables are created, the datasets can be built. I've included a timer here to see how long it takes to process:

```{r}
# Start the timer
t <- proc.time()

# Build palmsplus
palmsplus <- palms_build_palmsplus(palms)

# Build days
days <- palms_build_days(palmsplus)

# Build trajectories
trajectories <- palms_build_trajectories(palmsplus)

# Build multimodal trips
multimodal <- palms_build_multimodal(trajectories, 200, 10)

cat("Run time:", (proc.time() - t)[[1]]/60, "minutes")
```

Around 20 minutes for 191 participants isn't bad!

## Saving results

```{r, eval=FALSE}
write_csv(palmsplus, "palmsplus.csv")
write_csv(days, "days.csv")
write_csv(trajectories, "trajectories.csv")
write_csv(trajectories, "multimodal.csv")

st_write(palmsplus, "palmsplus.shp")
st_write(trajectories, "trajecories.shp")
st_write(trajectories, "multimodal.shp")
```


