---
title: "ggplot2 and palmsplusr"
author:
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4-ggplot2-palmsplusr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Install ggplot2

The development version of ggplot2 contains spatial plotting functions so it needs to be installed from Github. You will need to install devtools if you haven't already.

```{r, eval=FALSE}
devtools::install_github("tidyverse/ggplot2")
```


## Plotting 

The `geom_sf` is a new ggplot2 geom specifically for plotting simple features. You can plot points, lines, and polygons. Here is an example of looping through all participants, plotting their trajcetories, school and home on a map before saving it to file.

```{r, eval=FALSE}
library(ggplot2)

# Loop through all unique identifiers
for (i in unique(trajectories$identifier)) {
  
  # Get geometry for this person only
  t <- trajectories %>% filter(identifier == i)
  h <- home.buffer %>% filter(identifier == i)
  s <- school %>% filter(school_id == as.numeric(participant_basis[participant_basis$identifier == i, "school_id"]))

  # Plot the 3 geometry layers + add colours and title
  p <- ggplot() +
    geom_sf(data = t, aes(colour = factor(mot))) +
    geom_sf(data = h, aes(fill = "h"), alpha = 0.8) +
    geom_sf(data = s, aes(fill = "s"), alpha = 0.8) +
    scale_fill_manual(labels = c("Home (100m)", "Schoolyard"), 
                      values = c("#E01616", "#000080")) +
    scale_color_manual(breaks = c("1", "2", "3"), 
                       labels = c("Walk", "Bicycle", "Vehicle"),
                       values = c("1" = "forestgreen","2" = "#1874CD","3" = "#EE7600")) +
    theme_minimal() +
    labs(title = i, colour = "Trajectories", fill = "Polygons")

  # Save the plot to file
  ggsave(file = file.path("plots", paste0(i, ".png")), plot = p, width = 18, 
         height = 18, units = "cm")
}
```
