---
title: "A complete example"
author: Tom Stewart
date: "`r Sys.Date()`"
output: pdf_document
---

This article will demonstrate a real world example with a 191-participant dataset.

## Reading data

All data should be loaded at the beginning of the script. This includes the PALMS dataset, and any other shapefiles and csv files. Any preprocessing, such as buffering can also be done here.

```{r message=FALSE, warning=FALSE}
library(palmsplusr)
library(readr)

# Load PALMS dataset
palms <- read_palms("F:/data/csv/palms_output.csv")

# Load other csvs
participant_basis <- read_csv("F:/data/csv/participant_basis.csv")
class_timetable <- read_csv("F:/data/csv/class_timetable.csv")

# Load shapefiles (home points & schoolyard polygons)
home <- read_sf("F:/data/shapefiles/home.shp")
school <- read_sf("F:/data/shapefiles/school.shp")

# Buffer home points
home.buffer <- palms_buffer(point = home, distance = 100, crs = 2193)
```

\pagebreak

The participant_basis file is used to link participant identifier with the correct school and classroom: 
```{r}
head(participant_basis, n = 5)
```

The class_timetable file details start and end times for school and recess preiods, for each school/classroom:
```{r}
head(class_timetable, n = 5)
```

\pagebreak

## Creating the five field tables

All formulas for building `palmsplus`, `days`, `trajectories`, and `multimodal` are created. This is done by creating the five field tables:

* `palmsplus_fields`
* `palmsplus_domains`
* `trajectory_fields`
* `trajectory_locations`
* `multimodal_fields`

Most of these tables can be populated using the `palms_load_defaults()`. I use this function as a base, but then add additional fields that are specific to my dataset and research questions:

```{r}
# Remove all field tables that exist
palms_remove_tables()

# load defaults for:
# * palmsplus_fields
# * trajectory_fields
# * multimodal_fields
palms_load_defaults(palms_epoch(palms))

# palmsplus_fields (from other files)
palms_add_field("in_school_time", "palms_in_time(., i, class_timetable, 
                                   participant_basis, school_start, school_end)")
palms_add_field("in_recess1", "palms_in_time(., i, class_timetable, 
                               participant_basis, recess1_start, recess1_end)")
palms_add_field("in_recess2", "palms_in_time(., i, class_timetable, 
                               participant_basis, recess2_start, recess2_end)")
palms_add_field("at_home", "palms_in_polygon(., filter(home.buffer, identifier == i), identifier)")
palms_add_field("at_school", "palms_in_polygon(., filter(school, 
                              school_id == participant_basis %>% filter(identifier == i) %>% 
                              pull(school_id)))")

# palmsplus_domains
palms_add_domain("home", "at_home")
palms_add_domain("school", "(!at_home & at_school & in_school_time)")
palms_add_domain("transport", "!at_home & !(at_school & in_school_time) & 
                               (pedestrian | bicycle | vehicle)")
palms_add_domain("leisure", "!at_home & !(at_school & in_school_time) & 
                             !pedestrian & !bicycle & !vehicle")

# trajectory_locations
palms_add_trajectory_location("home_school", "at_home", "at_school")
palms_add_trajectory_location("school_home", "at_school", "at_home")
palms_add_trajectory_location("home_home", "at_home", "at_home")
palms_add_trajectory_location("school_school", "at_school", "at_school")
```

I didn't specify any additional `trajectory_fields` or `multimodal_fields` as the default fields will suffice. 

\pagebreak

All five of the field tables look like:

```{r, eval=FALSE}
palmsplus_fields
```
```{r echo=FALSE}
knitr::kable(palmsplus_fields, caption = "palmsplus_fields")
```

```{r, eval=FALSE}
palmsplus_domains
```
```{r echo=FALSE}
knitr::kable(palmsplus_domains, caption = "palmsplus_domains")
```

\pagebreak

```{r, eval=FALSE}
trajectory_fields
```
```{r echo=FALSE}
knitr::kable(trajectory_fields, caption = "trajectory_fields")
```

```{r, eval=FALSE}
trajectory_locations
```
```{r echo=FALSE}
knitr::kable(trajectory_locations, caption = "trajectory_locations")
```

```{r, eval=FALSE}
multimodal_fields
```
```{r echo=FALSE}
knitr::kable(multimodal_fields, caption = "multimodal_fields")
```

\pagebreak

## Building palmsplus, days, trajectories, and multimodal datasets

Once all of the formula tables are created, it is simple to build all of the datasets. I've included a timer here to see how long it takes to process:

```{r}
# Start the timer
t <- proc.time()

# Build palmsplus
palmsplus <- palms_build_palmsplus(palms, verbose = FALSE)

# Build days
days <- palms_build_days(palmsplus)

# Build trajectories
trajectories <- palms_build_trajectories(palmsplus)

# Build multimodal trips
multimodal <- palms_build_multimodal(trajectories, 200, 10)

cat("Run time:", (proc.time() - t)[[1]]/60, "minutes")
```

Just over 15 minutes for 191 participants isn't to bad!

## Saving results

```{r, eval=FALSE}
write_csv(palmsplus, "palmsplus.csv")
write_csv(days, "days.csv")
write_csv(trajectories, "trajectories.csv")
write_csv(multimodal, "multimodal.csv")

st_write(palmsplus, "palmsplus.shp")
st_write(trajectories, "trajecories.shp")
st_write(multimodal, "multimodal.shp")
```

